---
title: "Reproducible Analysis"
author:
  - Nascimento, M.^[Moyses Nascimento, Associate Professor, Department of Statistics - Federal University of Viçosa, moysesnascim@ufv.br]
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/nonlinear-ge-interaction-shap/
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
github-repo: wevertongomescosta/nonlinear-ge-interaction-shap
---

# Overview

This page presents the complete reproducible analytical workflow associated with the study **“Non-Linear Modeling of Genotype × Environment Interaction via Tree-Based Learning Algorithms and Shapley Decomposition.”**

The workflow combines phenotypic evaluation, Random Forest modeling, Shapley Additive Explanations, AMMI decomposition, stability indices, genotype ranking, and rank correlation.

All analytical code is presented directly in this R Markdown document and follows the order required to reproduce the analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%"
)
```

# Required packages

The following packages are used for data manipulation, graphical visualization, Random Forest modeling, SHAP decomposition, composite figure construction, genotype-rank visualization, and rank-correlation analysis.

```{r required-packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(randomForest)
library(DALEX)
library(patchwork)
library(ggrepel)
library(corrplot)
library(pals)
```

# Data import

The maize multi-environment trial data set is stored as a tab-delimited text file in the `data` directory.

```{r data-import}
maize <- read.table(
  "data/maize.txt",
  h = TRUE
)

head(maize)
```

# Data preparation

The original replicate-level observations are summarized by genotype and environment. The resulting data frame contains the mean grain yield for each genotype-by-environment combination.

```{r data-preparation}
df_ml <- maize %>%
  group_by(GEN, ENV) %>%
  summarise(
    GY = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml$ENV <- as.factor(df_ml$ENV)
df_ml$GEN <- as.factor(df_ml$GEN)
```

# Phenotypic performance

The distribution of mean grain yield is represented jointly across environments and genotypes.

Panel A displays the environmental distributions. Each jittered point represents the mean performance of one genotype within an environment. Panel B displays the marginal genotype distributions, with each point corresponding to the mean performance of one genotype in one environment.

The boxplot outliers are not displayed separately because all genotype-by-environment means are already represented by the jittered points.

```{r phenotypic-performance-figure, fig.width=12, fig.height=6, fig.align="center", warning=FALSE}
p_env <- ggplot(
  df_ml,
  aes(
    x = ENV,
    y = GY,
    fill = ENV
  )
) +
  geom_boxplot(
    alpha = 0.75,
    outlier.shape = NA,
    linewidth = 0.4
  ) +
  geom_jitter(
    width = 0.15,
    alpha = 0.4,
    color = "black",
    size = 1.2
  ) +
  scale_fill_manual(
    values = as.vector(
      pals::glasbey(20)
    )
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    x = "Environments",
    y = expression(
      paste(
        "Grain Yield (GY) in kg ",
        ha^-1
      )
    )
  ) +
  theme(
    legend.position = "none",
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    ),
    axis.text.x = element_text(
      color = "black",
      size = 9
    ),
    axis.text.y = element_text(
      color = "black",
      size = 9
    ),
    axis.title = element_text(
      size = 11
    )
  )

p_gen <- ggplot(
  df_ml,
  aes(
    x = GEN,
    y = GY,
    fill = GEN
  )
) +
  geom_boxplot(
    alpha = 0.75,
    outlier.shape = NA,
    linewidth = 0.4
  ) +
  geom_jitter(
    width = 0.15,
    alpha = 0.4,
    color = "black",
    size = 1.2
  ) +
  scale_fill_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    x = "Genotypes",
    y = expression(
      paste(
        "Grain Yield (GY) in kg ",
        ha^-1
      )
    )
  ) +
  theme(
    legend.position = "none",
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    ),
    axis.text.x = element_text(
      color = "black",
      size = 9
    ),
    axis.text.y = element_text(
      color = "black",
      size = 9
    ),
    axis.title = element_text(
      size = 11
    )
  )

figure_1 <- (
  p_env |
    p_gen
) +
  patchwork::plot_layout(
    widths = c(
      1.5,
      1
    )
  ) +
  patchwork::plot_annotation(
    tag_levels = "A"
  )

print(figure_1)
```

# Environmental index

The environmental index is calculated as the mean grain yield of all genotypes evaluated within each environment. This continuous variable represents the productive potential of each environment.

```{r environmental-index}
env_indices <- df_ml %>%
  group_by(ENV) %>%
  summarise(
    ENV_INDEX = mean(
      GY,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

df_ml <- df_ml %>%
  left_join(
    env_indices,
    by = "ENV"
  )

head(env_indices)
```

# Random Forest modeling

## Leave-one-environment-out validation

The Random Forest model is evaluated through leave-one-environment-out cross-validation. At each iteration, one environment is reserved for prediction, while the remaining environments are used for model fitting.

```{r random-forest-loeo, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(42)

ambientes <- unique(df_ml$ENV)
predicoes_cv <- data.frame()

cat("Starting leave-one-environment-out cross-validation...\n")

for (env_teste in ambientes) {
  
  dados_treino <- df_ml %>%
    filter(ENV != env_teste)
  
  dados_teste <- df_ml %>%
    filter(ENV == env_teste)
  
  rf_fold <- randomForest(
    GY ~ GEN + ENV_INDEX,
    data = dados_treino,
    ntree = 500
  )
  
  dados_teste$Pred_GY <- predict(
    rf_fold,
    newdata = dados_teste
  )
  
  predicoes_cv <- bind_rows(
    predicoes_cv,
    dados_teste
  )
}
```

```{r load-random-forest-validation, include=FALSE}
validation_file <- "output/random_forest_loeo_predictions.rds"

stopifnot(
  file.exists(validation_file)
)

predicoes_cv <- readRDS(
  validation_file
)

stopifnot(
  identical(
    dim(predicoes_cv),
    c(180L, 5L)
  ),
  identical(
    names(predicoes_cv),
    c(
      "GEN",
      "ENV",
      "GY",
      "ENV_INDEX",
      "Pred_GY"
    )
  ),
  nlevels(predicoes_cv$GEN) == 9L,
  nlevels(predicoes_cv$ENV) == 20L,
  !anyNA(predicoes_cv)
)
```

## Predictive metrics

Predictive performance is summarized using the cross-validated coefficient of determination, the Pearson correlation between observed and predicted grain yield, and the root mean squared error.

```{r predictive-metrics}
ss_res <- sum(
  (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
)

ss_tot <- sum(
  (predicoes_cv$GY - mean(predicoes_cv$GY))^2
)

r2_cv <- 1 - (ss_res / ss_tot)

rmse_cv <- sqrt(
  mean(
    (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
  )
)

accuracy_r <- cor(
  predicoes_cv$GY,
  predicoes_cv$Pred_GY
)

cat(
  "\n=== LEAVE-ONE-ENVIRONMENT-OUT VALIDATION METRICS ===\n"
)

cat(
  "Predictive capacity (cross-validated R²): ",
  round(r2_cv * 100, 2),
  "%\n"
)

cat(
  "Predictive accuracy (r):                  ",
  round(accuracy_r, 2),
  "\n"
)

cat(
  "Cross-validation RMSE:                   ",
  round(rmse_cv, 2),
  " kg ha-1\n"
)
```

## Final Random Forest model

The final Random Forest model is fitted using the complete analytical data set. Grain yield is modeled as a function of genotype and the continuous environmental index.

```{r final-random-forest, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(42)

rf_model <- randomForest(
  GY ~ GEN + ENV_INDEX,
  data = df_ml,
  ntree = 500
)
```

# SHAP decomposition

## Model explainer

The Random Forest model is connected to the DALEX framework using the genotype and environmental-index predictors. The response variable supplied to the explainer is the observed mean grain yield.

```{r dalex-explainer, eval=FALSE, message=FALSE, warning=FALSE}
explainer_rf <- explain(
  model = rf_model,
  data = df_ml[, c("GEN", "ENV_INDEX")],
  y = df_ml$GY,
  label = "Random Forest GxE (Continuous)"
)
```

## SHAP contribution values

SHAP contributions are calculated for every genotype-by-environment observation. For each observation, the mean contribution associated with genotype and environmental index is extracted from the SHAP decomposition.

```{r shap-contributions, eval=FALSE, message=FALSE, warning=FALSE}
shap_list <- list()

cat("Starting SHAP contribution calculation...\n")

for (i in 1:nrow(df_ml)) {
  
  obs <- df_ml[i, c("GEN", "ENV_INDEX")]
  
  shap_obs <- predict_parts(
    explainer = explainer_rf,
    new_observation = obs,
    type = "shap",
    B = 25
  )
  
  val_gen <- shap_obs %>%
    filter(variable_name == "GEN") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  val_env <- shap_obs %>%
    filter(variable_name == "ENV_INDEX") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  shap_list[[i]] <- data.frame(
    SHAP_GEN = val_gen,
    SHAP_ENV = val_env
  )
}

df_shap_final <- bind_rows(shap_list)

df_results <- cbind(
  df_ml,
  df_shap_final
)

head(df_results)
```

```{r load-shap-results, include=FALSE}
shap_file <- "output/shap_results.rds"

stopifnot(
  file.exists(shap_file)
)

df_results <- readRDS(
  shap_file
)

stopifnot(
  identical(
    dim(df_results),
    c(180L, 6L)
  ),
  identical(
    names(df_results),
    c(
      "GEN",
      "ENV",
      "GY",
      "ENV_INDEX",
      "SHAP_GEN",
      "SHAP_ENV"
    )
  ),
  nlevels(df_results$GEN) == 9L,
  nlevels(df_results$ENV) == 20L,
  !anyNA(df_results)
)
```

```{r shap-results-preview}
head(df_results)
```

## Genotype response profiles

Genotype-specific SHAP contributions are represented across environments. Positive and negative values indicate the direction of the genotype contribution to the Random Forest prediction relative to the model reference value.

```{r shap-genotype-profiles, fig.width=10, fig.height=7, fig.align="center", warning=FALSE}
df_results$GEN <- as.factor(df_results$GEN)
df_results$ENV <- as.factor(df_results$ENV)

p_shap <- ggplot(
  df_results,
  aes(
    x = ENV,
    y = SHAP_GEN,
    group = GEN,
    color = GEN
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_line(
    linewidth = 0.8,
    alpha = 0.85
  ) +
  geom_point(
    size = 2,
    alpha = 0.9
  ) +
  scale_color_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    title = " ",
    subtitle = " ",
    x = "Environments",
    y = bquote(
      "Genotype SHAP Value (G" %*% "E Effect) in kg " ~ ha^-1
    ),
    color = "Genotypes"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 13,
      hjust = 0
    ),
    plot.subtitle = element_text(
      size = 10,
      color = "gray30",
      margin = ggplot2::margin(
        b = 12
      )
    ),
    axis.title = element_text(
      face = "plain",
      size = 11
    ),
    axis.text = element_text(
      color = "black",
      size = 9
    ),
    legend.position = "bottom",
    legend.title = element_text(
      face = "bold",
      size = 10
    ),
    legend.text = element_text(
      size = 9
    ),
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    )
  )

print(p_shap)
```

# AMMI analysis

## Genotype-by-environment interaction matrix

The genotype-by-environment interaction effects are obtained after removing the genotype and environment main effects from the mean grain-yield observations.

```{r ammi-interaction-matrix}
grand_mean <- mean(
  df_results$GY,
  na.rm = TRUE
)

env_effects <- df_results %>%
  group_by(ENV) %>%
  summarise(
    m_env = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

gen_effects <- df_results %>%
  group_by(GEN) %>%
  summarise(
    m_gen = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_int <- df_results %>%
  left_join(
    env_effects,
    by = "ENV"
  ) %>%
  left_join(
    gen_effects,
    by = "GEN"
  ) %>%
  mutate(
    GE_effect = GY - m_env - m_gen + grand_mean
  )

mat_int <- df_int %>%
  select(
    GEN,
    ENV,
    GE_effect
  ) %>%
  pivot_wider(
    names_from = ENV,
    values_from = GE_effect
  ) %>%
  tibble::column_to_rownames("GEN") %>%
  as.matrix()
```

## Singular value decomposition

Singular value decomposition is applied to the genotype-by-environment interaction matrix. The squared singular values are used to calculate the percentage of interaction variation represented by each component.

```{r ammi-singular-value-decomposition}
svd_ammi <- svd(mat_int)

lambda <- svd_ammi$d

var_explained <- (
  lambda^2 / sum(lambda^2)
) * 100

ipca1_pct <- round(
  var_explained[1],
  2
)

ipca2_pct <- round(
  var_explained[2],
  2
)

data.frame(
  Component = paste0(
    "IPCA",
    seq_along(var_explained)
  ),
  Explained_variation = round(
    var_explained,
    2
  )
)
```

## Interaction principal components

The scores of the first two interaction principal components are obtained separately for genotypes and environments.

```{r ammi-interaction-scores}
scores_gen <- as.data.frame(
  svd_ammi$u[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_gen) <- c(
  "IPCA1",
  "IPCA2"
)

scores_gen$GEN <- as.factor(
  rownames(mat_int)
)

scores_env <- as.data.frame(
  svd_ammi$v[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_env) <- c(
  "IPCA1",
  "IPCA2"
)

scores_env$ENV <- as.factor(
  colnames(mat_int)
)
```

## AMMI biplot

The AMMI biplot represents genotype scores and environmental vectors for the first two interaction principal components. The percentages shown on the axes correspond to the interaction variation represented by each component.

```{r ammi-biplot, fig.width=10, fig.height=7, fig.align="center", warning=FALSE}
p_ammi <- ggplot() +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_segment(
    data = scores_env,
    aes(
      x = 0,
      y = 0,
      xend = IPCA1,
      yend = IPCA2
    ),
    arrow = arrow(
      length = grid::unit(
        0.18,
        "cm"
      )
    ),
    color = "gray30",
    alpha = 0.6,
    linewidth = 0.5
  ) +
  geom_text(
    data = scores_env,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = ENV
    ),
    vjust = 1.4,
    color = "gray20",
    size = 3
  ) +
  geom_point(
    data = scores_gen,
    aes(
      x = IPCA1,
      y = IPCA2,
      color = GEN
    ),
    size = 3.5,
    alpha = 0.9
  ) +
  geom_text(
    data = scores_gen,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = GEN
    ),
    vjust = -1.2,
    fontface = "bold",
    size = 3.5,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    x = paste0(
      "IPCA1 (",
      ipca1_pct,
      "%)"
    ),
    y = paste0(
      "IPCA2 (",
      ipca2_pct,
      "%)"
    ),
    color = "Genotypes"
  ) +
  theme(
    axis.title = element_text(
      face = "plain",
      size = 11
    ),
    axis.text = element_text(
      color = "black",
      size = 9
    ),
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    ),
    legend.position = "none"
  )

print(p_ammi)
```

# Joint AMMI and SHAP representation

The AMMI biplot and genotype-specific SHAP profiles are combined vertically. Panel A represents the first two interaction principal components, while Panel B represents the genotype SHAP contribution across environments.

```{r joint-ammi-shap, fig.width=10, fig.height=12, fig.align="center", warning=FALSE}
figura_final <- (
  p_ammi /
    p_shap
) +
  plot_layout(
    heights = c(
      1,
      1.1
    )
  ) +
  plot_annotation(
    tag_levels = "A"
  )

print(figura_final)
```

# Stability and selection indices

## AMMI Stability Value

The AMMI Stability Value combines the scores of the first two interaction principal components. Lower ASV values indicate greater stability according to the AMMI representation.

```{r ammi-stability-value}
peso_asv <- var_explained[1] / var_explained[2]

asv_gen <- scores_gen %>%
  mutate(
    ASV = sqrt(
      ((peso_asv * IPCA1)^2) +
        (IPCA2^2)
    )
  ) %>%
  select(
    GEN,
    ASV
  )

asv_gen
```

## SHAP effect and stability

The mean genotype SHAP contribution represents the average contribution of each genotype across environments. The standard deviation of the genotype SHAP contributions represents the variation of this contribution across environments.

```{r shap-effect-stability}
tab_shap <- df_results %>%
  group_by(GEN) %>%
  summarise(
    Media_GY = mean(GY),
    SHAP_Medio = mean(SHAP_GEN),
    SHAP_SD = sd(SHAP_GEN),
    .groups = "drop"
  )

tab_shap
```

## Simultaneous selection indices

The classical Simultaneous Selection Index combines the grain-yield ranking and the AMMI Stability Value ranking through their rank sum.

A SHAP-based Simultaneous Selection Index is also calculated. Mean genotype SHAP contribution and SHAP stability are first transformed to a common scale ranging from 0 to 100. Higher standardized values indicate more favorable performance.

The SHAP-based index assigns a weight of 60% to the standardized mean SHAP contribution and 40% to the standardized SHAP stability component.

```{r simultaneous-selection-index}
theta1 <- 60
theta2 <- 40

tabela_comparativa <- left_join(
  tab_shap,
  asv_gen,
  by = "GEN"
)

dados_resultado <- tabela_comparativa %>%
  mutate(
    Rank_Prod = rank(
      -Media_GY,
      ties.method = "first"
    ),
    Rank_ASV = rank(
      ASV,
      ties.method = "first"
    ),
    Rank_SHAP_Eff = rank(
      -SHAP_Medio,
      ties.method = "first"
    ),
    Rank_SHAP_Stb = rank(
      SHAP_SD,
      ties.method = "first"
    ),
    SHAP_Medio_scaled = 100 * (
      SHAP_Medio - min(SHAP_Medio)
    ) / (
      max(SHAP_Medio) - min(SHAP_Medio)
    ),
    SHAP_SD_scaled = 100 * (
      max(SHAP_SD) - SHAP_SD
    ) / (
      max(SHAP_SD) - min(SHAP_SD)
    ),
    SHAP_SSI = (
      SHAP_Medio_scaled * theta1 +
        SHAP_SD_scaled * theta2
    ) / (
      theta1 + theta2
    ),
    Rank_SHAP_SSI = rank(
      -SHAP_SSI,
      ties.method = "first"
    ),
    Rank_SSI = rank(
      Rank_Prod + Rank_ASV,
      ties.method = "first"
    )
  ) %>%
  arrange(Rank_Prod)
```

# Genotype ranking

Genotypes are compared according to grain yield, AMMI stability, the classical simultaneous selection index, mean SHAP contribution, SHAP stability, and the SHAP-based simultaneous selection index. In every ranking, position 1 represents the preferred genotype according to the corresponding criterion.

```{r genotype-ranking}
print(dados_resultado)
```

# Rank dynamics

The rank-dynamics plot represents how the relative position of each genotype changes across grain yield, AMMI stability, simultaneous selection, SHAP stability, and SHAP effect. Rank 1 represents the preferred position under every criterion.

```{r rank-dynamics-data}
dados_long <- dados_resultado %>%
  pivot_longer(
    cols = c(
      Rank_Prod,
      Rank_ASV,
      Rank_SSI,
      Rank_SHAP_Stb,
      Rank_SHAP_SSI,
      Rank_SHAP_Eff
    ),
    names_to = "Metodologia",
    values_to = "Posicao"
  ) %>%
  mutate(
    Metodologia = factor(
      Metodologia,
      levels = c(
        "Rank_Prod",
        "Rank_ASV",
        "Rank_SSI",
        "Rank_SHAP_Stb",
        "Rank_SHAP_SSI",
        "Rank_SHAP_Eff"
      ),
      labels = c(
        "Grain Yield",
        "ASV (Linear)",
        "SSI (Linear)",
        "SHAP Stability",
        "SHAP SSI",
        "SHAP Effect"
      )
    )
  )
```

```{r rank-dynamics-plot, fig.width=12, fig.height=7, fig.align="center", warning=FALSE}
figure_3_bump <- ggplot(
  dados_long,
  aes(
    x = Metodologia,
    y = Posicao,
    group = GEN,
    color = GEN
  )
) +
  geom_line(
    linewidth = 1.2,
    alpha = 0.8
  ) +
  geom_point(
    size = 4,
    shape = 21,
    fill = "white",
    stroke = 1.5
  ) +
  geom_text(
    data = subset(
      dados_long,
      Metodologia == "Grain Yield"
    ),
    aes(label = GEN),
    hjust = 1.5,
    fontface = "bold",
    size = 4
  ) +
  geom_text(
    data = subset(
      dados_long,
      Metodologia == "SHAP Effect"
    ),
    aes(label = GEN),
    hjust = -0.5,
    fontface = "bold",
    size = 4
  ) +
  scale_color_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  scale_y_reverse(
    breaks = 1:9
  ) +
  theme_classic(
    base_size = 13
  ) +
  labs(
    x = "Selection Framework",
    y = "Genotype Rank (1 = Best)"
  ) +
  theme(
    legend.position = "none",
    axis.line = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.y = element_line(
      color = "gray93",
      linetype = "dashed"
    ),
    panel.grid.major.x = element_line(
      color = "gray95",
      linewidth = 8
    ),
    plot.margin = ggplot2::margin(
      t = 10,
      r = 25,
      b = 10,
      l = 25
    )
  )

print(figure_3_bump)
```

# Stability concordance

The concordance plot compares the AMMI Stability Value ranking, which represents linear interaction stability, with the SHAP stability ranking, which represents variation in the non-linear genotype contribution across environments.

Each point represents one genotype. Lower ranking values indicate greater stability. Therefore, genotypes in the lower-left region are classified as relatively stable by both approaches. The dotted identity line represents exact agreement between the two stability rankings, while the point fill represents the grain-yield ranking.

```{r stability-concordance, fig.width=9, fig.height=7, fig.align="center", warning=FALSE}
p_stability_concordance <- ggplot(
  dados_resultado,
  aes(
    x = Rank_ASV,
    y = Rank_SHAP_Stb,
    label = GEN
  )
) +
  geom_vline(
    xintercept = 5,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.5
  ) +
  geom_hline(
    yintercept = 5,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.5
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dotted",
    color = "red",
    alpha = 0.6,
    linewidth = 0.7
  ) +
  geom_point(
    aes(fill = Rank_Prod),
    size = 5.5,
    shape = 21,
    color = "black",
    stroke = 0.8
  ) +
  scale_fill_viridis_c(
    option = "mako",
    direction = -1,
    breaks = c(
      1,
      5,
      9
    ),
    labels = c(
      "High (1)",
      "Medium (5)",
      "Low (9)"
    )
  ) +
  ggrepel::geom_text_repel(
    fontface = "bold",
    size = 4,
    box.padding = 0.4,
    point.padding = 0.3
  ) +
  scale_x_continuous(
    breaks = 1:9,
    limits = c(
      0.5,
      9.5
    )
  ) +
  scale_y_continuous(
    breaks = 1:9,
    limits = c(
      0.5,
      9.5
    )
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    x = "Linear Stability Ranking (Rank_ASV)",
    y = "Non-Linear Stability Ranking (Rank_SHAP_Stb)",
    fill = "Grain Yield"
  ) +
  theme(
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    ),
    axis.text = element_text(
      color = "black"
    ),
    legend.position = "right",
    legend.title = element_text(
      face = "bold",
      size = 10
    )
  )

print(p_stability_concordance)
```

# Spearman rank correlation

Spearman rank correlations are calculated among the six genotype-selection criteria. Positive correlations indicate that two criteria tend to assign similar positions to the genotypes, whereas negative correlations indicate opposing ranking patterns.

The comparison includes grain yield, AMMI stability, classical simultaneous selection, SHAP stability, SHAP-based simultaneous selection, and mean SHAP effect.

## Correlation matrix

```{r spearman-correlation-matrix}
colunas_rank <- dados_resultado %>%
  select(
    Rank_Prod,
    Rank_ASV,
    Rank_SSI,
    Rank_SHAP_Stb,
    Rank_SHAP_SSI,
    Rank_SHAP_Eff
  ) %>%
  as.data.frame()

colnames(colunas_rank) <- c(
  "Yield",
  "ASV (Linear)",
  "SSI (Linear)",
  "SHAP Stability",
  "SHAP SSI",
  "SHAP Effect"
)

matriz_corr <- cor(
  colunas_rank,
  method = "spearman"
)

round(
  matriz_corr,
  2
)
```

## Graphical representation

The upper triangular correlation matrix represents the direction and magnitude of the Spearman associations. The numerical coefficients are displayed inside the corresponding cells.

```{r spearman-correlation-plot, fig.width=10, fig.height=8, fig.align="center", warning=FALSE}
rampa_customizada_artigo <- colorRampPalette(
  c(
    "#B2182B",
    "#E08214",
    "#FDB863",
    "#F7F7F7",
    "#B2ABD2",
    "#8073AC",
    "#542788"
  )
)(200)

corrplot::corrplot(
  matriz_corr,
  method = "color",
  type = "upper",
  col = rampa_customizada_artigo,
  addCoef.col = "black",
  number.cex = 1.25,
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 1.25,
  cl.cex = 1.1,
  diag = FALSE,
  tl.pos = "td",
  mar = c(
    1,
    1,
    4,
    1
  )
)
```
