---
title: "Reproducible Analysis"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Overview

This page presents the complete reproducible analytical workflow associated with the study **“Non-Linear Modeling of Genotype × Environment Interaction via Tree-Based Learning Algorithms and Shapley Decomposition.”**

The workflow combines phenotypic evaluation, Random Forest modeling, Shapley Additive Explanations, AMMI decomposition, stability indices, genotype ranking, and rank correlation.

All analytical code is presented directly in this R Markdown document and follows the order required to reproduce the analysis.

```{r setup, eval=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = "100%",
  fig.height = 6
)
```

# Required packages

The following packages are used for data manipulation, graphical visualization, Random Forest modeling, SHAP decomposition, and composite figure construction.

```{r required-packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(randomForest)
library(DALEX)
library(patchwork)
```

# Data import

The maize multi-environment trial data set is stored as a tab-delimited text file in the `data` directory.

```{r data-import}
maize <- read.table(
  "data/maize.txt",
  h = TRUE
)

head(maize)
```

# Data preparation

The original replicate-level observations are summarized by genotype and environment. The resulting data frame contains the mean grain yield for each genotype-by-environment combination.

```{r data-preparation}
df_ml <- maize %>%
  group_by(GEN, ENV) %>%
  summarise(
    GY = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml$ENV <- as.factor(df_ml$ENV)
df_ml$GEN <- as.factor(df_ml$GEN)
```

# Phenotypic performance

## Grain yield across environments

The distribution of mean grain yield is represented for each environment. Each point corresponds to the mean performance of one genotype within an environment.

```{r grain-yield-environments, fig.width=10, fig.height=6, fig.align="center", warning=FALSE}
ggplot(df_ml, aes(x = ENV, y = GY, fill = ENV)) +
  geom_boxplot(
    alpha = 0.7,
    show.legend = FALSE
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.5,
    color = "black"
  ) +
  labs(
    title = "",
    x = "Environment (ENV)",
    y = "Grain Yield (GY)"
  ) +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(
    legend.position = "none"
  )
```

## Grain yield across genotypes

The distribution of mean grain yield is represented for each genotype. Each point corresponds to the mean performance of one genotype within an environment.

```{r grain-yield-genotypes, fig.width=10, fig.height=6, fig.align="center", warning=FALSE}
ggplot(df_ml, aes(x = GEN, y = GY, fill = GEN)) +
  geom_boxplot(
    alpha = 0.7
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.5,
    color = "black"
  ) +
  labs(
    title = "",
    x = "Genotype (GEN)",
    y = "Grain Yield (GY)"
  ) +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(
    legend.position = "none"
  )
```

# Environmental index

The environmental index is calculated as the mean grain yield of all genotypes evaluated within each environment. This continuous variable represents the productive potential of each environment.

```{r environmental-index}
env_indices <- df_ml %>%
  group_by(ENV) %>%
  summarise(
    ENV_INDEX = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml <- df_ml %>%
  left_join(env_indices, by = "ENV")

head(env_indices)
```

# Random Forest modeling

## Leave-one-environment-out validation

The Random Forest model is evaluated through leave-one-environment-out cross-validation. At each iteration, one environment is reserved for prediction, while the remaining environments are used for model fitting.

```{r random-forest-loeo, message=FALSE, warning=FALSE}
set.seed(42)

ambientes <- unique(df_ml$ENV)
predicoes_cv <- data.frame()

cat("Starting leave-one-environment-out cross-validation...\n")

for (env_teste in ambientes) {
  
  dados_treino <- df_ml %>%
    filter(ENV != env_teste)
  
  dados_teste <- df_ml %>%
    filter(ENV == env_teste)
  
  rf_fold <- randomForest(
    GY ~ GEN + ENV_INDEX,
    data = dados_treino,
    ntree = 500
  )
  
  dados_teste$Pred_GY <- predict(
    rf_fold,
    newdata = dados_teste
  )
  
  predicoes_cv <- bind_rows(
    predicoes_cv,
    dados_teste
  )
}
```

## Predictive metrics

Predictive performance is summarized using the cross-validated coefficient of determination, the Pearson correlation between observed and predicted grain yield, and the root mean squared error.

```{r predictive-metrics}
ss_res <- sum(
  (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
)

ss_tot <- sum(
  (predicoes_cv$GY - mean(predicoes_cv$GY))^2
)

r2_cv <- 1 - (ss_res / ss_tot)

rmse_cv <- sqrt(
  mean(
    (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
  )
)

accuracy_r <- cor(
  predicoes_cv$GY,
  predicoes_cv$Pred_GY
)

cat(
  "\n=== LEAVE-ONE-ENVIRONMENT-OUT VALIDATION METRICS ===\n"
)

cat(
  "Predictive capacity (cross-validated R²): ",
  round(r2_cv * 100, 2),
  "%\n"
)

cat(
  "Predictive accuracy (r):                  ",
  round(accuracy_r, 2),
  "\n"
)

cat(
  "Cross-validation RMSE:                   ",
  round(rmse_cv, 2),
  " kg ha-1\n"
)
```

## Final Random Forest model

The final Random Forest model is fitted using the complete analytical data set. Grain yield is modeled as a function of genotype and the continuous environmental index.

```{r final-random-forest, message=FALSE, warning=FALSE}
set.seed(42)

rf_model <- randomForest(
  GY ~ GEN + ENV_INDEX,
  data = df_ml,
  ntree = 500
)
```

# SHAP decomposition

## Model explainer

The Random Forest model is connected to the DALEX framework using the genotype and environmental-index predictors. The response variable supplied to the explainer is the observed mean grain yield.

```{r dalex-explainer, message=FALSE, warning=FALSE}
explainer_rf <- explain(
  model = rf_model,
  data = df_ml[, c("GEN", "ENV_INDEX")],
  y = df_ml$GY,
  label = "Random Forest GxE (Continuous)"
)
```

## SHAP contribution values

SHAP contributions are calculated for every genotype-by-environment observation. For each observation, the mean contribution associated with genotype and environmental index is extracted from the SHAP decomposition.

```{r shap-contributions, message=FALSE, warning=FALSE}
shap_list <- list()

cat("Starting SHAP contribution calculation...\n")

for (i in 1:nrow(df_ml)) {
  
  obs <- df_ml[i, c("GEN", "ENV_INDEX")]
  
  shap_obs <- predict_parts(
    explainer = explainer_rf,
    new_observation = obs,
    type = "shap",
    B = 25
  )
  
  val_gen <- shap_obs %>%
    filter(variable_name == "GEN") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  val_env <- shap_obs %>%
    filter(variable_name == "ENV_INDEX") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  shap_list[[i]] <- data.frame(
    SHAP_GEN = val_gen,
    SHAP_ENV = val_env
  )
}

df_shap_final <- bind_rows(shap_list)

df_results <- cbind(
  df_ml,
  df_shap_final
)

head(df_results)
```

## Genotype response profiles

Genotype-specific SHAP contributions are represented across environments. Positive and negative values indicate the direction of the genotype contribution to the Random Forest prediction relative to the model reference value.

```{r shap-genotype-profiles, fig.width=10, fig.height=7, fig.align="center", warning=FALSE}
df_results$GEN <- as.factor(df_results$GEN)
df_results$ENV <- as.factor(df_results$ENV)

p_shap <- ggplot(
  df_results,
  aes(
    x = ENV,
    y = SHAP_GEN,
    group = GEN,
    color = GEN
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_line(
    linewidth = 0.8,
    alpha = 0.85
  ) +
  geom_point(
    size = 2,
    alpha = 0.9
  ) +
  scale_color_viridis_d(
    option = "viridis"
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    title = "",
    subtitle = "",
    x = "Environments",
    y = expression(
      paste(
        "Genotype SHAP Value (G",
        times,
        "E Effect) in kg ",
        ha^-1
      )
    ),
    color = "Genotypes"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 13,
      hjust = 0
    ),
    plot.subtitle = element_text(
      size = 10,
      color = "gray30",
      margin = ggplot2::margin(b = 12)
    ),
    axis.title = element_text(
      face = "plain",
      size = 11
    ),
    axis.text = element_text(
      color = "black",
      size = 9
    ),
    legend.position = "bottom",
    legend.title = element_text(
      face = "bold",
      size = 10
    ),
    legend.text = element_text(
      size = 9
    ),
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    )
  )

print(p_shap)
```

# AMMI analysis

## Genotype-by-environment interaction matrix

The genotype-by-environment interaction effects are obtained after removing the genotype and environment main effects from the mean grain-yield observations.

```{r ammi-interaction-matrix}
grand_mean <- mean(
  df_results$GY,
  na.rm = TRUE
)

env_effects <- df_results %>%
  group_by(ENV) %>%
  summarise(
    m_env = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

gen_effects <- df_results %>%
  group_by(GEN) %>%
  summarise(
    m_gen = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_int <- df_results %>%
  left_join(
    env_effects,
    by = "ENV"
  ) %>%
  left_join(
    gen_effects,
    by = "GEN"
  ) %>%
  mutate(
    GE_effect = GY - m_env - m_gen + grand_mean
  )

mat_int <- df_int %>%
  select(
    GEN,
    ENV,
    GE_effect
  ) %>%
  pivot_wider(
    names_from = ENV,
    values_from = GE_effect
  ) %>%
  tibble::column_to_rownames("GEN") %>%
  as.matrix()
```

## Singular value decomposition

Singular value decomposition is applied to the genotype-by-environment interaction matrix. The squared singular values are used to calculate the percentage of interaction variation represented by each component.

```{r ammi-singular-value-decomposition}
svd_ammi <- svd(mat_int)

lambda <- svd_ammi$d

var_explained <- (
  lambda^2 / sum(lambda^2)
) * 100

ipca1_pct <- round(
  var_explained[1],
  2
)

ipca2_pct <- round(
  var_explained[2],
  2
)

data.frame(
  Component = paste0(
    "IPCA",
    seq_along(var_explained)
  ),
  Explained_variation = round(
    var_explained,
    2
  )
)
```

## Interaction principal components

The scores of the first two interaction principal components are obtained separately for genotypes and environments.

```{r ammi-interaction-scores}
scores_gen <- as.data.frame(
  svd_ammi$u[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_gen) <- c(
  "IPCA1",
  "IPCA2"
)

scores_gen$GEN <- as.factor(
  rownames(mat_int)
)

scores_env <- as.data.frame(
  svd_ammi$v[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_env) <- c(
  "IPCA1",
  "IPCA2"
)

scores_env$ENV <- as.factor(
  colnames(mat_int)
)
```

## AMMI biplot

The AMMI biplot represents genotype scores and environmental vectors for the first two interaction principal components. The percentages shown on the axes correspond to the interaction variation represented by each component.

```{r ammi-biplot, fig.width=10, fig.height=7, fig.align="center", warning=FALSE}
p_ammi <- ggplot() +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_segment(
    data = scores_env,
    aes(
      x = 0,
      y = 0,
      xend = IPCA1,
      yend = IPCA2
    ),
    arrow = arrow(
      length = grid::unit(
        0.18,
        "cm"
      )
    ),
    color = "gray30",
    alpha = 0.6,
    linewidth = 0.5
  ) +
  geom_text(
    data = scores_env,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = ENV
    ),
    vjust = 1.4,
    color = "gray20",
    size = 3
  ) +
  geom_point(
    data = scores_gen,
    aes(
      x = IPCA1,
      y = IPCA2,
      color = GEN
    ),
    size = 3.5,
    alpha = 0.9
  ) +
  geom_text(
    data = scores_gen,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = GEN
    ),
    vjust = -1.2,
    fontface = "bold",
    size = 3.5,
    show.legend = FALSE
  ) +
  scale_color_viridis_d(
    option = "viridis"
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    x = paste0(
      "IPCA1 (",
      ipca1_pct,
      "%)"
    ),
    y = paste0(
      "IPCA2 (",
      ipca2_pct,
      "%)"
    ),
    color = "Genotypes"
  ) +
  theme(
    axis.title = element_text(
      face = "plain",
      size = 11
    ),
    axis.text = element_text(
      color = "black",
      size = 9
    ),
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    ),
    legend.position = "none"
  )

print(p_ammi)
```

# Joint AMMI and SHAP representation

The AMMI biplot and genotype-specific SHAP profiles are combined vertically. Panel A represents the first two interaction principal components, while Panel B represents the genotype SHAP contribution across environments.

```{r joint-ammi-shap, fig.width=10, fig.height=12, fig.align="center", warning=FALSE}
figura_final <- (
  p_ammi /
    p_shap
) +
  plot_layout(
    heights = c(
      1,
      1.1
    )
  ) +
  plot_annotation(
    tag_levels = "A"
  )

print(figura_final)
```

# Stability and selection indices

## AMMI Stability Value

The AMMI Stability Value is calculated from the first two interaction principal components.

## SHAP effect and stability

Mean genotype SHAP contribution and the standard deviation of genotype SHAP contributions are summarized for each genotype.

## Simultaneous Selection Index

The Simultaneous Selection Index combines grain-yield and AMMI stability rankings.

# Genotype ranking

Genotypes are compared according to:

- grain yield;
- AMMI Stability Value;
- Simultaneous Selection Index;
- SHAP effect;
- SHAP stability.

# Rank dynamics

A bump chart represents changes in genotype positions across the evaluated criteria.

# Stability concordance

The relationship between AMMI stability ranking and SHAP stability ranking is represented using a rank-concordance plot.

# Spearman rank correlation

Spearman correlations are calculated among grain yield, AMMI stability, simultaneous selection, SHAP effect, and SHAP stability rankings.