---
title: "Reproducible Analysis"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Overview

This page presents the complete reproducible analytical workflow associated with the study **“Non-Linear Modeling of Genotype × Environment Interaction via Tree-Based Learning Algorithms and Shapley Decomposition.”**

The workflow combines phenotypic evaluation, Random Forest modeling, Shapley Additive Explanations, AMMI decomposition, stability indices, genotype ranking, and rank correlation.

All analytical code is presented directly in this R Markdown document and follows the order required to reproduce the analysis.

```{r setup, eval=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = "100%",
  fig.height = 6
)
```

# Required packages

The following packages are used for data manipulation, graphical visualization, Random Forest modeling, and SHAP decomposition.

```{r required-packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(randomForest)
library(DALEX)
```

# Data import

The maize multi-environment trial data set is stored as a tab-delimited text file in the `data` directory.

```{r data-import}
maize <- read.table(
  "data/maize.txt",
  h = TRUE
)

head(maize)
```

# Data preparation

The original replicate-level observations are summarized by genotype and environment. The resulting data frame contains the mean grain yield for each genotype-by-environment combination.

```{r data-preparation}
df_ml <- maize %>%
  group_by(GEN, ENV) %>%
  summarise(
    GY = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml$ENV <- as.factor(df_ml$ENV)
df_ml$GEN <- as.factor(df_ml$GEN)
```

# Phenotypic performance

## Grain yield across environments

The distribution of mean grain yield is represented for each environment. Each point corresponds to the mean performance of one genotype within an environment.

```{r grain-yield-environments, fig.width=10, fig.height=6, fig.align="center", warning=FALSE}
ggplot(df_ml, aes(x = ENV, y = GY, fill = ENV)) +
  geom_boxplot(
    alpha = 0.7,
    show.legend = FALSE
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.5,
    color = "black"
  ) +
  labs(
    title = "",
    x = "Environment (ENV)",
    y = "Grain Yield (GY)"
  ) +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(
    legend.position = "none"
  )
```

## Grain yield across genotypes

The distribution of mean grain yield is represented for each genotype. Each point corresponds to the mean performance of one genotype within an environment.

```{r grain-yield-genotypes, fig.width=10, fig.height=6, fig.align="center", warning=FALSE}
ggplot(df_ml, aes(x = GEN, y = GY, fill = GEN)) +
  geom_boxplot(
    alpha = 0.7
  ) +
  geom_jitter(
    width = 0.2,
    alpha = 0.5,
    color = "black"
  ) +
  labs(
    title = "",
    x = "Genotype (GEN)",
    y = "Grain Yield (GY)"
  ) +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(
    legend.position = "none"
  )
```

# Environmental index

The environmental index is calculated as the mean grain yield of all genotypes evaluated within each environment. This continuous variable represents the productive potential of each environment.

```{r environmental-index}
env_indices <- df_ml %>%
  group_by(ENV) %>%
  summarise(
    ENV_INDEX = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml <- df_ml %>%
  left_join(env_indices, by = "ENV")

head(env_indices)
```

# Random Forest modeling

## Leave-one-environment-out validation

The Random Forest model is evaluated through leave-one-environment-out cross-validation. At each iteration, one environment is reserved for prediction, while the remaining environments are used for model fitting.

```{r random-forest-loeo, message=FALSE, warning=FALSE}
set.seed(42)

ambientes <- unique(df_ml$ENV)
predicoes_cv <- data.frame()

cat("Starting leave-one-environment-out cross-validation...\n")

for (env_teste in ambientes) {
  
  dados_treino <- df_ml %>%
    filter(ENV != env_teste)
  
  dados_teste <- df_ml %>%
    filter(ENV == env_teste)
  
  rf_fold <- randomForest(
    GY ~ GEN + ENV_INDEX,
    data = dados_treino,
    ntree = 500
  )
  
  dados_teste$Pred_GY <- predict(
    rf_fold,
    newdata = dados_teste
  )
  
  predicoes_cv <- bind_rows(
    predicoes_cv,
    dados_teste
  )
}
```

## Predictive metrics

Predictive performance is summarized using the cross-validated coefficient of determination, the Pearson correlation between observed and predicted grain yield, and the root mean squared error.

```{r predictive-metrics}
ss_res <- sum(
  (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
)

ss_tot <- sum(
  (predicoes_cv$GY - mean(predicoes_cv$GY))^2
)

r2_cv <- 1 - (ss_res / ss_tot)

rmse_cv <- sqrt(
  mean(
    (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
  )
)

accuracy_r <- cor(
  predicoes_cv$GY,
  predicoes_cv$Pred_GY
)

cat(
  "\n=== LEAVE-ONE-ENVIRONMENT-OUT VALIDATION METRICS ===\n"
)

cat(
  "Predictive capacity (cross-validated R²): ",
  round(r2_cv * 100, 2),
  "%\n"
)

cat(
  "Predictive accuracy (r):                  ",
  round(accuracy_r, 2),
  "\n"
)

cat(
  "Cross-validation RMSE:                   ",
  round(rmse_cv, 2),
  " kg ha-1\n"
)
```

## Final Random Forest model

The final Random Forest model is fitted using the complete analytical data set. Grain yield is modeled as a function of genotype and the continuous environmental index.

```{r final-random-forest, message=FALSE, warning=FALSE}
set.seed(42)

rf_model <- randomForest(
  GY ~ GEN + ENV_INDEX,
  data = df_ml,
  ntree = 500
)
```

# SHAP decomposition

## Model explainer

The Random Forest model is connected to the DALEX framework using the genotype and environmental-index predictors. The response variable supplied to the explainer is the observed mean grain yield.

```{r dalex-explainer, message=FALSE, warning=FALSE}
explainer_rf <- explain(
  model = rf_model,
  data = df_ml[, c("GEN", "ENV_INDEX")],
  y = df_ml$GY,
  label = "Random Forest GxE (Continuous)"
)
```

## SHAP contribution values

SHAP contributions are calculated for every genotype-by-environment observation. For each observation, the mean contribution associated with genotype and environmental index is extracted from the SHAP decomposition.

```{r shap-contributions, message=FALSE, warning=FALSE}
shap_list <- list()

cat("Starting SHAP contribution calculation...\n")

for (i in 1:nrow(df_ml)) {
  
  obs <- df_ml[i, c("GEN", "ENV_INDEX")]
  
  shap_obs <- predict_parts(
    explainer = explainer_rf,
    new_observation = obs,
    type = "shap",
    B = 25
  )
  
  val_gen <- shap_obs %>%
    filter(variable_name == "GEN") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  val_env <- shap_obs %>%
    filter(variable_name == "ENV_INDEX") %>%
    summarise(mean(contribution)) %>%
    pull()
  
  shap_list[[i]] <- data.frame(
    SHAP_GEN = val_gen,
    SHAP_ENV = val_env
  )
}

df_shap_final <- bind_rows(shap_list)

df_results <- cbind(
  df_ml,
  df_shap_final
)

head(df_results)
```

## Genotype response profiles

Genotype-specific SHAP contributions are represented across environments. Positive and negative values indicate the direction of the genotype contribution to the Random Forest prediction relative to the model reference value.

```{r shap-genotype-profiles, fig.width=10, fig.height=7, fig.align="center", warning=FALSE}
df_results$GEN <- as.factor(df_results$GEN)
df_results$ENV <- as.factor(df_results$ENV)

p_shap <- ggplot(
  df_results,
  aes(
    x = ENV,
    y = SHAP_GEN,
    group = GEN,
    color = GEN
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_line(
    linewidth = 0.8,
    alpha = 0.85
  ) +
  geom_point(
    size = 2,
    alpha = 0.9
  ) +
  scale_color_viridis_d(
    option = "viridis"
  ) +
  theme_classic(
    base_size = 12
  ) +
  labs(
    title = "",
    subtitle = "",
    x = "Environments",
    y = expression(
      paste(
        "Genotype SHAP Value (G",
        times,
        "E Effect) in kg ",
        ha^-1
      )
    ),
    color = "Genotypes"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 13,
      hjust = 0
    ),
    plot.subtitle = element_text(
      size = 10,
      color = "gray30",
      margin = ggplot2::margin(b = 12)
    ),
    axis.title = element_text(
      face = "plain",
      size = 11
    ),
    axis.text = element_text(
      color = "black",
      size = 9
    ),
    legend.position = "bottom",
    legend.title = element_text(
      face = "bold",
      size = 10
    ),
    legend.text = element_text(
      size = 9
    ),
    axis.line = element_line(
      color = "black",
      linewidth = 0.5
    )
  )

print(p_shap)
```

# AMMI analysis

## Genotype-by-environment interaction matrix

The interaction matrix is obtained after removing genotype and environment main effects.

## Singular value decomposition

Singular value decomposition is applied to the genotype-by-environment interaction matrix.

## Interaction principal components

The first two interaction principal components are extracted for genotypes and environments.

## AMMI biplot

The AMMI biplot represents genotype scores and environmental vectors for the first two interaction principal components.

# Joint AMMI and SHAP representation

The AMMI biplot and genotype SHAP profiles are combined into a single graphical representation.

# Stability and selection indices

## AMMI Stability Value

The AMMI Stability Value is calculated from the first two interaction principal components.

## SHAP effect and stability

Mean genotype SHAP contribution and the standard deviation of genotype SHAP contributions are summarized for each genotype.

## Simultaneous Selection Index

The Simultaneous Selection Index combines grain-yield and AMMI stability rankings.

# Genotype ranking

Genotypes are compared according to:

- grain yield;
- AMMI Stability Value;
- Simultaneous Selection Index;
- SHAP effect;
- SHAP stability.

# Rank dynamics

A bump chart represents changes in genotype positions across the evaluated criteria.

# Stability concordance

The relationship between AMMI stability ranking and SHAP stability ranking is represented using a rank-concordance plot.

# Spearman rank correlation

Spearman correlations are calculated among grain yield, AMMI stability, simultaneous selection, SHAP effect, and SHAP stability rankings.