---
title: "Reproducible Analysis"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Overview

This page presents the complete reproducible analytical workflow associated with the study **“Non-Linear Modeling of Genotype × Environment Interaction via Tree-Based Learning Algorithms and Shapley Decomposition.”**

The workflow combines phenotypic evaluation, Random Forest modeling, Shapley Additive Explanations, AMMI decomposition, stability indices, genotype ranking, and rank correlation.

All analytical code is presented directly in this R Markdown document and follows the order required to reproduce the analysis.

# Computational setup

The global chunk options define how code, messages, warnings, tables, and figures are displayed throughout the analysis.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 6
)
```

# Required packages

This section loads the packages required for data manipulation, machine learning, model interpretation, statistical analysis, and graphical visualization.

# Data import

This section imports the original maize multi-environment trial data set stored in:

```text
data/maize.txt
```

# Data preparation

The original observations are summarized by genotype and environment before the machine-learning and interaction analyses.

# Phenotypic performance

## Grain yield across environments

This section presents the distribution of grain yield across the evaluated environments.

## Grain yield across genotypes

This section presents the distribution of grain yield across the evaluated genotypes.

# Environmental index

The environmental index represents the mean grain-yield performance observed in each environment.

# Random Forest modeling

## Leave-one-environment-out validation

The Random Forest model is evaluated through leave-one-environment-out cross-validation.

## Predictive metrics

Predictive performance is summarized using:

- coefficient of determination;
- Pearson correlation between observed and predicted values;
- root mean squared error.

## Final Random Forest model

The final Random Forest model is fitted using the complete analytical data set.

# SHAP decomposition

## Model explainer

A model explainer is created using the DALEX framework.

## SHAP contribution values

SHAP values are obtained for the genotype and environmental-index predictors.

## Genotype response profiles

Genotype-specific SHAP contributions are represented across environments.

# AMMI analysis

## Genotype-by-environment interaction matrix

The interaction matrix is obtained after removing genotype and environment main effects.

## Singular value decomposition

Singular value decomposition is applied to the genotype-by-environment interaction matrix.

## Interaction principal components

The first two interaction principal components are extracted for genotypes and environments.

## AMMI biplot

The AMMI biplot represents genotype scores and environmental vectors for the first two interaction principal components.

# Joint AMMI and SHAP representation

The AMMI biplot and genotype SHAP profiles are combined into a single graphical representation.

# Stability and selection indices

## AMMI Stability Value

The AMMI Stability Value is calculated from the first two interaction principal components.

## SHAP effect and stability

Mean genotype SHAP contribution and the standard deviation of genotype SHAP contributions are summarized for each genotype.

## Simultaneous Selection Index

The Simultaneous Selection Index combines grain-yield and AMMI stability rankings.

# Genotype ranking

Genotypes are compared according to:

- grain yield;
- AMMI Stability Value;
- Simultaneous Selection Index;
- SHAP effect;
- SHAP stability.

# Rank dynamics

A bump chart represents changes in genotype positions across the evaluated criteria.

# Stability concordance

The relationship between AMMI stability ranking and SHAP stability ranking is represented using a rank-concordance plot.

# Spearman rank correlation

Spearman correlations are calculated among grain yield, AMMI stability, simultaneous selection, SHAP effect, and SHAP stability rankings.

# Session information

The computational environment used to build the analysis is reported below.

```{r session-information}
sessionInfo()
```