---
title: "Reproducible Analysis"
author:
  - Nascimento, M.^[Moyses Nascimento, Associate Professor, Department of Statistics - Federal University of Viçosa, moysesnascim@ufv.br]
  - Costa, W. G.^[Weverton Gomes da Costa, Postdoctoral Researcher, Department of Statistics - Federal University of Viçosa, weverton.costa@ufv.br]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/nonlinear-ge-interaction-shap/
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
github-repo: wevertongomescosta/nonlinear-ge-interaction-shap
---

# Overview

This page presents the complete reproducible analytical workflow associated with the study **“Non-Linear Modeling of Genotype × Environment Interaction via Tree-Based Learning Algorithms and Shapley Decomposition.”**

The workflow combines phenotypic evaluation, Random Forest modeling, Shapley Additive Explanations, AMMI decomposition, stability indices, genotype ranking, and rank correlation.

All analytical code is presented directly in this R Markdown document and follows the order required to reproduce the analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%"
)
```

# Required packages

The following packages are used for data manipulation, graphical visualization, Random Forest modeling, SHAP decomposition, faceted scale control, genotype-rank visualization, and rank-correlation analysis.

```{r required-packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(randomForest)
library(DALEX)
library(iBreakDown)
library(pals)
library(ggh4x)
library(ggcorrplot)
```

# Data import

The maize multi-environment trial data set is stored as a tab-delimited text file in the `data` directory.

```{r data-import}
maize <- read.table(
  "data/maize.txt",
  h = TRUE
)

head(maize)
```

# Data preparation

The original replicate-level observations are summarized by genotype and environment. The resulting data frame contains the mean grain yield for each genotype-by-environment combination.

```{r data-preparation}
df_ml <- maize %>%
  group_by(GEN, ENV) %>%
  summarise(
    GY = mean(GY, na.rm = TRUE),
    .groups = "drop"
  )

df_ml$ENV <- as.factor(df_ml$ENV)
df_ml$GEN <- as.factor(df_ml$GEN)
```

# Leave-one-environment-out validation scheme

The leave-one-environment-out design uses each environment once as the validation environment. The diagonal cells identify the environment excluded in each iteration, while the remaining environments are retained for model fitting.

```{r loeo-validation-scheme, fig.width=7, fig.height=5, fig.align="center", warning=FALSE}
loeo_scheme <- tidyr::expand_grid(
  Iteration_number = seq_len(20),
  Environment_number = seq_len(20)
) %>%
  mutate(
    Value = as.numeric(
      Iteration_number == Environment_number
    ),
    Iteration = factor(
      paste0("Iter", Iteration_number),
      levels = rev(
        paste0("Iter", seq_len(20))
      )
    ),
    Environment = factor(
      paste0("Env", Environment_number),
      levels = paste0(
        "Env",
        seq_len(20)
      )
    )
  )

figure_1 <- ggplot(
  loeo_scheme,
  aes(
    x = Environment,
    y = Iteration,
    fill = Value
  )
) +
  geom_tile(
    color = "white",
    linewidth = 0.1,
    alpha = 0.5
  ) +
  scale_fill_gradient(
    low = "#4ADC36",
    high = "#DB241F",
    guide = "none"
  ) +
  labs(
    x = "Environments",
    y = "Iterations"
  ) +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = ggplot2::margin(
      t = 2,
      r = 0,
      b = 0,
      l = 0
    ),
    axis.line = element_blank(),
    axis.text.y = element_text(
      size = 7,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.text.x = element_text(
      size = 7,
      angle = 90,
      vjust = 0.5,
      hjust = 0.5,
      color = "black",
      family = "serif"
    ),
    axis.title.y = element_text(
      size = 12,
      color = "black",
      family = "serif"
    ),
    axis.title.x = element_text(
      size = 10,
      color = "black",
      family = "serif"
    ),
    axis.ticks = element_blank()
  )

print(figure_1)
```

# Phenotypic performance

The distribution of mean grain yield is represented in two vertically arranged panels. Panel A displays the environmental distributions, and Panel B displays the genotype distributions. Every jittered point represents one genotype-by-environment mean.

The boxplot outliers are not displayed separately because all genotype-by-environment means are already represented by the jittered points.

```{r phenotypic-performance-figure, fig.width=8, fig.height=8, fig.align="center", warning=FALSE}
phenotypic_plot_data <- bind_rows(
  df_ml %>%
    transmute(
      facet = "A",
      Category = factor(
        as.character(ENV),
        levels = as.character(seq_len(20))
      ),
      GY = GY
    ),
  df_ml %>%
    transmute(
      facet = "B",
      Category = factor(
        as.character(GEN),
        levels = as.character(seq_len(20))
      ),
      GY = GY
    )
)

phenotypic_titles <- data.frame(
  facet = c("A", "B"),
  Category = factor(
    c("10", "5"),
    levels = as.character(seq_len(20))
  ),
  x_title = c(
    "Environments",
    "Genotypes"
  )
)

figure_2 <- ggplot(
  phenotypic_plot_data,
  aes(
    x = Category,
    y = GY,
    fill = Category
  )
) +
  facet_wrap(
    ~facet,
    scales = "free",
    ncol = 1
  ) +
  geom_boxplot(
    alpha = 0.75,
    outlier.shape = NA,
    linewidth = 0.4
  ) +
  geom_jitter(
    width = 0.15,
    alpha = 0.4,
    color = "black",
    size = 1.2
  ) +
  geom_text(
    data = phenotypic_titles,
    aes(
      x = Category,
      y = -Inf,
      label = x_title
    ),
    inherit.aes = FALSE,
    vjust = 3,
    size = 3.5,
    family = "serif",
    color = "black"
  ) +
  scale_fill_manual(
    values = stats::setNames(
      as.vector(
        pals::glasbey(20)
      ),
      as.character(seq_len(20))
    )
  ) +
  coord_cartesian(
    clip = "off"
  ) +
  labs(
    x = "",
    y = expression(
      paste(
        "Grain Yield (GY) in kg ",
        ha^-1
      )
    )
  ) +
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(
      color = "black",
      fill = "white"
    ),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = ggplot2::margin(
      t = 2,
      r = 0,
      b = 8,
      l = 0
    ),
    axis.line.x = element_blank(),
    axis.line.y = element_line(
      color = "black"
    ),
    axis.text.y = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.text.x = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.title.y = element_text(
      size = 12,
      color = "black",
      family = "serif"
    ),
    axis.title.x = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(
      hjust = 0,
      size = 9,
      family = "serif",
      color = "black"
    )
  )

print(figure_2)
```

# Environmental index

The environmental index is calculated as the mean grain yield of all genotypes evaluated within each environment. This continuous variable represents the productive potential of each environment.

```{r environmental-index}
env_indices <- df_ml %>%
  group_by(ENV) %>%
  summarise(
    ENV_INDEX = mean(
      GY,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

df_ml <- df_ml %>%
  left_join(
    env_indices,
    by = "ENV"
  )

head(env_indices)
```

# Random Forest modeling

## Leave-one-environment-out validation

The Random Forest model is evaluated through leave-one-environment-out cross-validation. At each iteration, one environment is reserved for prediction, while the remaining environments are used for model fitting.

```{r random-forest-loeo, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(42)

ambientes <- unique(df_ml$ENV)
predicoes_cv <- data.frame()

cat("Starting leave-one-environment-out cross-validation...\n")

for (env_teste in ambientes) {
  dados_treino <- df_ml %>%
    filter(
      ENV != env_teste
    )

  dados_teste <- df_ml %>%
    filter(
      ENV == env_teste
    )

  rf_fold <- randomForest(
    GY ~ GEN + ENV_INDEX,
    data = dados_treino,
    ntree = 500
  )

  dados_teste$Pred_GY <- predict(
    rf_fold,
    newdata = dados_teste
  )

  predicoes_cv <- bind_rows(
    predicoes_cv,
    dados_teste
  )
}

predicoes_cv <- predicoes_cv %>%
  arrange(
    ENV,
    GEN
  )

validation_file <- "output/random_forest_loeo_predictions.rds"

dir.create(
  dirname(validation_file),
  recursive = TRUE,
  showWarnings = FALSE
)

saveRDS(
  predicoes_cv,
  validation_file
)
```

```{r load-random-forest-validation, include=FALSE}
validation_file <- "output/random_forest_loeo_predictions.rds"

if (!exists(
  "predicoes_cv",
  inherits = FALSE
)) {
  stopifnot(
    file.exists(validation_file)
  )

  predicoes_cv <- readRDS(
    validation_file
  )
}

stopifnot(
  identical(
    dim(predicoes_cv),
    c(180L, 5L)
  ),
  identical(
    names(predicoes_cv),
    c(
      "GEN",
      "ENV",
      "GY",
      "ENV_INDEX",
      "Pred_GY"
    )
  ),
  nlevels(predicoes_cv$GEN) == 9L,
  nlevels(predicoes_cv$ENV) == 20L,
  !anyNA(predicoes_cv)
)
```

## Predictive metrics

Predictive performance is summarized using the cross-validated coefficient of determination, the Pearson correlation between observed and predicted grain yield, and the root mean squared error.

```{r predictive-metrics}
ss_res <- sum(
  (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
)

ss_tot <- sum(
  (predicoes_cv$GY - mean(predicoes_cv$GY))^2
)

r2_cv <- 1 - (ss_res / ss_tot)

rmse_cv <- sqrt(
  mean(
    (predicoes_cv$GY - predicoes_cv$Pred_GY)^2
  )
)

accuracy_r <- cor(
  predicoes_cv$GY,
  predicoes_cv$Pred_GY
)

cat(
  "\n=== LEAVE-ONE-ENVIRONMENT-OUT VALIDATION METRICS ===\n"
)

cat(
  "Predictive capacity (cross-validated R²): ",
  round(r2_cv * 100, 2),
  "%\n"
)

cat(
  "Predictive accuracy (r):                  ",
  round(accuracy_r, 2),
  "\n"
)

cat(
  "Cross-validation RMSE:                   ",
  round(rmse_cv, 2),
  " kg ha-1\n"
)
```

## Final Random Forest model

The final Random Forest model is fitted using the complete analytical data set. Grain yield is modeled as a function of genotype and the continuous environmental index.

```{r final-random-forest, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(42)

rf_model <- randomForest(
  GY ~ GEN + ENV_INDEX,
  data = df_ml,
  ntree = 500
)
```

# SHAP decomposition

## Model explainer

The Random Forest model is connected to the DALEX framework using the genotype and environmental-index predictors. The response variable supplied to the explainer is the observed mean grain yield.

```{r dalex-explainer, eval=FALSE, message=FALSE, warning=FALSE}
explainer_rf <- explain(
  model = rf_model,
  data = df_ml[
    ,
    c(
      "GEN",
      "ENV_INDEX"
    )
  ],
  y = df_ml$GY,
  label = "Random Forest GxE (Continuous)"
)
```

## SHAP contribution values

SHAP contributions are calculated for every genotype-by-environment observation. For each observation, the mean contribution associated with genotype and environmental index is extracted from the SHAP decomposition.

```{r shap-contributions, eval=FALSE, message=FALSE, warning=FALSE}
set.seed(42)

shap_list <- vector(
  mode = "list",
  length = nrow(df_ml)
)

cat("Starting SHAP contribution calculation...\n")

for (i in seq_len(nrow(df_ml))) {
  obs <- df_ml[
    i,
    c(
      "GEN",
      "ENV_INDEX"
    )
  ]

  shap_obs <- predict_parts(
    explainer = explainer_rf,
    new_observation = obs,
    type = "shap",
    B = 25
  )

  val_gen <- shap_obs %>%
    filter(
      variable_name == "GEN"
    ) %>%
    summarise(
      SHAP_GEN = mean(contribution),
      .groups = "drop"
    ) %>%
    pull(SHAP_GEN)

  val_env <- shap_obs %>%
    filter(
      variable_name == "ENV_INDEX"
    ) %>%
    summarise(
      SHAP_ENV = mean(contribution),
      .groups = "drop"
    ) %>%
    pull(SHAP_ENV)

  shap_list[[i]] <- data.frame(
    SHAP_GEN = val_gen,
    SHAP_ENV = val_env
  )
}

df_shap_final <- bind_rows(
  shap_list
)

df_results <- bind_cols(
  df_ml,
  df_shap_final
) %>%
  arrange(
    ENV,
    GEN
  )

shap_file <- "output/shap_results.rds"

dir.create(
  dirname(shap_file),
  recursive = TRUE,
  showWarnings = FALSE
)

saveRDS(
  df_results,
  shap_file
)
```

```{r load-shap-results, include=FALSE}
shap_file <- "output/shap_results.rds"

if (!exists(
  "df_results",
  inherits = FALSE
)) {
  stopifnot(
    file.exists(shap_file)
  )

  df_results <- readRDS(
    shap_file
  )
}

stopifnot(
  identical(
    dim(df_results),
    c(180L, 6L)
  ),
  identical(
    names(df_results),
    c(
      "GEN",
      "ENV",
      "GY",
      "ENV_INDEX",
      "SHAP_GEN",
      "SHAP_ENV"
    )
  ),
  nlevels(df_results$GEN) == 9L,
  nlevels(df_results$ENV) == 20L,
  !anyNA(df_results)
)
```

```{r shap-results-preview}
head(df_results)
```

# AMMI analysis

## Genotype-by-environment interaction matrix

The genotype-by-environment interaction effects are obtained after removing the genotype and environment main effects from the mean grain-yield observations.

```{r ammi-interaction-matrix}
grand_mean <- mean(
  df_results$GY,
  na.rm = TRUE
)

env_effects <- df_results %>%
  group_by(ENV) %>%
  summarise(
    m_env = mean(
      GY,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

gen_effects <- df_results %>%
  group_by(GEN) %>%
  summarise(
    m_gen = mean(
      GY,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

df_int <- df_results %>%
  left_join(
    env_effects,
    by = "ENV"
  ) %>%
  left_join(
    gen_effects,
    by = "GEN"
  ) %>%
  mutate(
    GE_effect = GY - m_env - m_gen + grand_mean
  )

mat_int <- df_int %>%
  select(
    GEN,
    ENV,
    GE_effect
  ) %>%
  pivot_wider(
    names_from = ENV,
    values_from = GE_effect
  ) %>%
  tibble::column_to_rownames(
    "GEN"
  ) %>%
  as.matrix()
```

## Singular value decomposition

Singular value decomposition is applied to the genotype-by-environment interaction matrix. The squared singular values are used to calculate the percentage of interaction variation represented by each component.

```{r ammi-singular-value-decomposition}
svd_ammi <- svd(
  mat_int
)

lambda <- svd_ammi$d

var_explained <- (
  lambda^2 / sum(lambda^2)
) * 100

ipca1_pct <- round(
  var_explained[1],
  2
)

ipca2_pct <- round(
  var_explained[2],
  2
)

data.frame(
  Component = paste0(
    "IPCA",
    seq_along(var_explained)
  ),
  Explained_variation = round(
    var_explained,
    2
  )
)
```

## Interaction principal components

The scores of the first two interaction principal components are obtained separately for genotypes and environments.

```{r ammi-interaction-scores}
scores_gen <- as.data.frame(
  svd_ammi$u[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_gen) <- c(
  "IPCA1",
  "IPCA2"
)

scores_gen$GEN <- as.factor(
  rownames(mat_int)
)

scores_env <- as.data.frame(
  svd_ammi$v[, 1:2] %*%
    diag(
      sqrt(lambda[1:2])
    )
)

colnames(scores_env) <- c(
  "IPCA1",
  "IPCA2"
)

scores_env$ENV <- as.factor(
  colnames(mat_int)
)
```

# Joint AMMI and SHAP representation

The first panel represents the genotype scores and environmental vectors for the first two AMMI interaction principal components. The second panel represents genotype-specific SHAP contributions across environments.

```{r joint-ammi-shap, fig.width=8, fig.height=10, fig.align="center", warning=FALSE}
ammi_gen_plot_data <- scores_gen %>%
  mutate(
    facet = "A"
  )

ammi_env_plot_data <- scores_env %>%
  mutate(
    facet = "A"
  )

shap_plot_data <- df_results %>%
  transmute(
    GEN = as.factor(GEN),
    ENV = as.numeric(
      as.character(ENV)
    ),
    SHAP_GEN = SHAP_GEN,
    facet = "B"
  )

figure_3_titles <- data.frame(
  facet = c(
    "A",
    "B"
  ),
  x_title = c(
    paste0(
      "IPCA1 (",
      ipca1_pct,
      "%)"
    ),
    "Environments"
  ),
  y_title = c(
    paste0(
      "IPCA2 (",
      ipca2_pct,
      "%)"
    ),
    "Genotype SHAP Value (G × E Effect) in kg ha⁻¹"
  ),
  x = c(
    -7,
    10
  ),
  y = c(
    -7,
    -100
  )
)

figure_3 <- ggplot() +
  facet_wrap(
    ~facet,
    scales = "free",
    ncol = 1
  ) +
  coord_cartesian(
    clip = "off"
  ) +
  geom_vline(
    data = data.frame(
      facet = "A",
      x = 0
    ),
    aes(
      xintercept = x
    ),
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_hline(
    data = data.frame(
      facet = c(
        "A",
        "B"
      ),
      y = 0
    ),
    aes(
      yintercept = y
    ),
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.4
  ) +
  geom_segment(
    data = ammi_env_plot_data,
    aes(
      x = 0,
      y = 0,
      xend = IPCA1,
      yend = IPCA2
    ),
    arrow = arrow(
      length = grid::unit(
        0.18,
        "cm"
      )
    ),
    color = "gray30",
    alpha = 0.6,
    linewidth = 0.5
  ) +
  geom_text(
    data = ammi_env_plot_data,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = ENV
    ),
    vjust = 1.4,
    family = "serif",
    color = "gray20",
    size = 3
  ) +
  geom_point(
    data = ammi_gen_plot_data,
    aes(
      x = IPCA1,
      y = IPCA2,
      color = GEN
    ),
    size = 3.5,
    alpha = 0.9
  ) +
  geom_text(
    data = ammi_gen_plot_data,
    aes(
      x = IPCA1,
      y = IPCA2,
      label = GEN
    ),
    vjust = -1.2,
    family = "serif",
    fontface = "bold",
    size = 3.5,
    show.legend = FALSE
  ) +
  geom_line(
    data = shap_plot_data,
    aes(
      x = ENV,
      y = SHAP_GEN,
      group = GEN,
      color = GEN
    ),
    linewidth = 0.5
  ) +
  geom_point(
    data = shap_plot_data,
    aes(
      x = ENV,
      y = SHAP_GEN,
      group = GEN,
      color = GEN
    ),
    size = 1
  ) +
  geom_text(
    data = figure_3_titles,
    aes(
      x = x,
      y = -Inf,
      label = x_title
    ),
    inherit.aes = FALSE,
    vjust = 4.5,
    size = 3,
    family = "serif",
    color = "black"
  ) +
  geom_text(
    data = figure_3_titles,
    aes(
      x = -Inf,
      y = y,
      label = y_title
    ),
    inherit.aes = FALSE,
    vjust = -5,
    size = 3,
    family = "serif",
    color = "black",
    angle = 90
  ) +
  scale_color_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  labs(
    x = "",
    y = "",
    color = "Genotype"
  ) +
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(
      color = "black",
      fill = "white"
    ),
    panel.grid = element_blank(),
    legend.position = "bottom",
    plot.margin = ggplot2::margin(
      t = 2,
      r = 0,
      b = 0,
      l = 0
    ),
    legend.text = element_text(
      size = 10,
      family = "serif"
    ),
    legend.title = element_text(
      size = 10,
      family = "serif",
      vjust = 0.5
    ),
    axis.line.x = element_blank(),
    axis.line.y = element_line(
      color = "black"
    ),
    axis.text.y = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.text.x = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.key = element_blank(),
    strip.text = element_text(
      hjust = 0,
      size = 9,
      family = "serif",
      color = "black"
    )
  ) +
  ggh4x::facetted_pos_scales(
    x = list(
      facet == "A" ~ scale_x_continuous(
        breaks = seq(
          -50,
          50,
          20
        )
      ),
      facet == "B" ~ scale_x_continuous(
        breaks = seq_len(20)
      )
    ),
    y = list(
      facet == "A" ~ scale_y_continuous(
        limits = c(
          -40,
          30
        )
      ),
      facet == "B" ~ scale_y_continuous()
    )
  )

print(figure_3)
```

# Stability and selection indices

## AMMI Stability Value

The AMMI Stability Value combines the scores of the first two interaction principal components. Lower ASV values indicate greater stability according to the AMMI representation.

```{r ammi-stability-value}
peso_asv <- var_explained[1] /
  var_explained[2]

asv_gen <- scores_gen %>%
  mutate(
    ASV = sqrt(
      ((peso_asv * IPCA1)^2) +
        (IPCA2^2)
    )
  ) %>%
  select(
    GEN,
    ASV
  )

asv_gen
```

## SHAP effect and stability

The mean genotype SHAP contribution represents the average contribution of each genotype across environments. The standard deviation of the genotype SHAP contributions represents the variation of this contribution across environments.

```{r shap-effect-stability}
tab_shap <- df_results %>%
  group_by(GEN) %>%
  summarise(
    Media_GY = mean(GY),
    SHAP_Medio = mean(SHAP_GEN),
    SHAP_SD = sd(SHAP_GEN),
    .groups = "drop"
  )

tab_shap
```

## Simultaneous selection indices

The classical Simultaneous Selection Index combines the grain-yield ranking and the AMMI Stability Value ranking through their rank sum.

The SHAP-based Simultaneous Selection Index combines the mean genotype SHAP contribution and SHAP stability after min–max normalization to a common scale from 0 to 100. Higher normalized values indicate more favorable performance. The index assigns a weight of 60% to mean SHAP contribution and 40% to SHAP stability.

```{r simultaneous-selection-index}
theta1 <- 60
theta2 <- 40

dados_resultado <- left_join(
  tab_shap,
  asv_gen,
  by = "GEN"
) %>%
  mutate(
    Rank_Prod = rank(
      -Media_GY,
      ties.method = "first"
    ),
    Rank_ASV = rank(
      ASV,
      ties.method = "first"
    ),
    Rank_SHAP_Eff = rank(
      -SHAP_Medio,
      ties.method = "first"
    ),
    Rank_SHAP_Stb = rank(
      SHAP_SD,
      ties.method = "first"
    ),
    SHAP_Medio_scaled = 100 * (
      SHAP_Medio - min(SHAP_Medio)
    ) / (
      max(SHAP_Medio) - min(SHAP_Medio)
    ),
    SHAP_SD_scaled = 100 * (
      max(SHAP_SD) - SHAP_SD
    ) / (
      max(SHAP_SD) - min(SHAP_SD)
    ),
    SHAP_SSI = (
      SHAP_Medio_scaled * theta1 +
        SHAP_SD_scaled * theta2
    ) / (
      theta1 + theta2
    ),
    Rank_SHAP_SSI = rank(
      -SHAP_SSI,
      ties.method = "first"
    ),
    Rank_SSI = rank(
      Rank_Prod + Rank_ASV,
      ties.method = "first"
    )
  ) %>%
  arrange(
    Rank_Prod
  )
```

# Genotype ranking

Genotypes are compared according to grain yield, AMMI stability, the classical simultaneous selection index, mean SHAP contribution, SHAP stability, and the SHAP-based simultaneous selection index. In every ranking, position 1 represents the preferred genotype according to the corresponding criterion.

```{r genotype-ranking}
print(
  dados_resultado
)
```

# Rank dynamics

The rank-dynamics plot represents how the relative position of each genotype changes across grain yield, AMMI stability, classical simultaneous selection, SHAP stability, SHAP-based simultaneous selection, and mean SHAP effect. Rank 1 represents the preferred position under every criterion.

```{r rank-dynamics-data}
dados_long <- dados_resultado %>%
  pivot_longer(
    cols = c(
      Rank_Prod,
      Rank_ASV,
      Rank_SSI,
      Rank_SHAP_Stb,
      Rank_SHAP_SSI,
      Rank_SHAP_Eff
    ),
    names_to = "Metodologia",
    values_to = "Posicao"
  ) %>%
  mutate(
    Metodologia = factor(
      Metodologia,
      levels = c(
        "Rank_Prod",
        "Rank_ASV",
        "Rank_SSI",
        "Rank_SHAP_Stb",
        "Rank_SHAP_SSI",
        "Rank_SHAP_Eff"
      ),
      labels = c(
        "Grain Yield",
        "ASV (Linear)",
        "SSI (Linear)",
        "SHAP Stability",
        "SHAP SSI",
        "SHAP Effect"
      )
    )
  )
```

```{r rank-dynamics-plot, fig.width=8, fig.height=6, fig.align="center", warning=FALSE}
figure_4 <- ggplot(
  dados_long,
  aes(
    x = Metodologia,
    y = Posicao,
    group = GEN,
    color = GEN
  )
) +
  geom_line(
    linewidth = 1,
    alpha = 0.8
  ) +
  geom_point(
    size = 2,
    shape = 21,
    fill = "white",
    stroke = 1.5
  ) +
  geom_text(
    data = subset(
      dados_long,
      Metodologia == "Grain Yield"
    ),
    aes(
      label = GEN
    ),
    hjust = 2,
    fontface = "bold",
    size = 4,
    family = "serif"
  ) +
  geom_text(
    data = subset(
      dados_long,
      Metodologia == "SHAP Effect"
    ),
    aes(
      label = GEN
    ),
    hjust = -1,
    fontface = "bold",
    size = 4,
    family = "serif"
  ) +
  scale_color_manual(
    values = as.vector(
      pals::glasbey(9)
    )
  ) +
  scale_y_reverse(
    breaks = 1:9
  ) +
  coord_cartesian(
    clip = "off"
  ) +
  labs(
    x = "Selection Framework",
    y = "Genotype Rank (1 = Best)"
  ) +
  theme_classic(
    base_size = 13
  ) +
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(
      color = "black",
      fill = "white"
    ),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = ggplot2::margin(
      t = 2,
      r = 18,
      b = 2,
      l = 18
    ),
    axis.line.x = element_blank(),
    axis.line.y = element_line(
      color = "black"
    ),
    axis.text.y = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.text.x = element_text(
      size = 10,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.title.y = element_text(
      size = 12,
      color = "black",
      family = "serif"
    ),
    axis.title.x = element_text(
      size = 10,
      color = "black",
      family = "serif"
    ),
    axis.ticks = element_blank()
  )

print(figure_4)
```

# Spearman rank correlation

Spearman rank correlations are calculated among the six genotype-selection criteria. Positive correlations indicate that two criteria tend to assign similar positions to the genotypes, whereas negative correlations indicate opposing ranking patterns.

The comparison includes grain yield, AMMI stability, classical simultaneous selection, SHAP stability, SHAP-based simultaneous selection, and mean SHAP effect.

## Correlation matrix

```{r spearman-correlation-matrix}
colunas_rank <- dados_resultado %>%
  select(
    Rank_Prod,
    Rank_ASV,
    Rank_SSI,
    Rank_SHAP_Stb,
    Rank_SHAP_SSI,
    Rank_SHAP_Eff
  ) %>%
  as.data.frame()

colnames(
  colunas_rank
) <- c(
  "Yield",
  "ASV (Linear)",
  "SSI (Linear)",
  "SHAP Stability",
  "SHAP SSI",
  "SHAP Effect"
)

matriz_corr <- cor(
  colunas_rank,
  method = "spearman"
)

round(
  matriz_corr,
  2
)
```

## Graphical representation

The upper triangular correlation matrix represents the direction and magnitude of the Spearman associations. The numerical coefficients are displayed inside the corresponding cells.

```{r spearman-correlation-plot, fig.width=8, fig.height=7, fig.align="center", warning=FALSE}
figure_5 <- ggcorrplot::ggcorrplot(
  matriz_corr,
  method = "square",
  lab = TRUE,
  lab_size = 4,
  show.diag = FALSE,
  type = "upper",
  legend.title = "Correlation",
  colors = c(
    "#DB241F",
    "white",
    "#4ADC36"
  )
) +
  labs(
    x = "",
    y = ""
  ) +
  theme(
    strip.background = element_rect(
      color = "black",
      fill = "white"
    ),
    panel.background = element_rect(
      color = "black",
      fill = "white"
    ),
    panel.grid = element_blank(),
    legend.position = "right",
    plot.margin = ggplot2::margin(
      t = 0,
      r = 0,
      b = 0,
      l = 0
    ),
    legend.text = element_text(
      size = 12,
      family = "serif"
    ),
    legend.title = element_text(
      size = 14,
      family = "serif",
      vjust = 0.5
    ),
    axis.line.x = element_blank(),
    axis.line.y = element_line(
      color = "black"
    ),
    axis.text.y = element_text(
      size = 12,
      angle = 0,
      color = "black",
      family = "serif"
    ),
    axis.text.x = element_text(
      size = 12,
      angle = 90,
      hjust = 0.5,
      vjust = 0.5,
      color = "black",
      family = "serif"
    ),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.key = element_blank(),
    strip.text = element_text(
      size = 14,
      family = "serif",
      color = "black"
    )
  )

print(figure_5)
```
